Multiple imputation of incomplete multilevel data using Heckman selection models

### 2022-10-24 1.Paper/Paper.Rmd
update text

---
title: Multiple imputation of incomplete multilevel data using Heckman selection models
author:
- name: J. Munoz*
  num: a
- name: V. de Jong
  num: a
- name: T. Debray
  num: a
address:
- num: a
  org: Julius Center for Health Sciences and Primary Care, Utrecht Medical Center,Utrech, The Netherlands
corres: "* Johanna Munoz. \\email{j.munozavila@umcutrecht.nl}"
presentaddress: Julius Center for Health Sciences and Primary Care, Utrecht Medical Center, Universiteitsweg 100, 3584 CG Utrech, The Netherlands
authormark: Uthor \emph{et al}.
articletype: Research article
received: 2022-11-01
revised: 2022-12-01
accepted: 2023-01-01
abstract: "Missing data is a common problem in medical research, and is commonly addressed using multiple imputation. Although traditional imputation methods allow for valid statistical inference when data are missing at random (MAR), their implementation is not justified when observations are clustered (e.g., within studies) or when the presence of missingness depends on unobserved information. Unfortunately, this situation is increasingly common, and typically arises when individual participant data (IPD) are combined from multiple studies. While several imputation methods have been proposed for addressing individual studies where data are missing not at random (MNAR), their application and validity in large datasets with clustering remains unclear. We therefore explored the consequence of MNAR data in IPDMA in-depth, and proposed novel multilevel imputation methods for common missing patterns in clustered datasets. These methods build upon the principles of Heckman selection models, and adopt a two-stage meta-analysis approach for imputing binary and continuous variables. After evaluating the proposed imputation models in simulated scenarios, we illustrated their use in a real IPDMA based on a malaria study in five subregions in Uganda, where we focused on estimating the prevalence of parasitemia for children aged 2-10 years."
keywords: Heckman model; IPDMA; Missing not at random; Selection models;Multiple imputation;
bibliography: bibfile.bib
output: 
  rticles::sim_article:
  fig_caption: yes
  longtable: yes
  doublespace: yes

@@

Under the BVN assumption, the parameters of the Heckman model coefficients can be estimated using the two-step Heckman method (H2S) [@heckman1976] or the full information maximum likelihood method (FIML) [@amemiya1984]. However, both methods can lead to inconsistent estimators when this underlying distribution cannot be assumed[@gomes_etal20].To overcome this problem, other approaches have been proposed that relax the distribution assumptions, among them copula models [@smith03]. The copula approach uses a function, known as copula, that joins the marginal distribution of the error terms of the selection and outcome equation which are specified separately.
$$ F(r^*_{i},y^*_{i}) =\Phi\left(r^*_{i}-X^S_{i}\beta^S_{i},\frac{y^*_{i}-X^O_{i}\beta^O_{i}}{\sigma_i}, \rho\right)$$
Thus, to estimate the parameters of the Heckman method, it is sufficient to specify the marginal distributions of the error terms, and link them with a suitable copula function. In our imputation method we estimate the Heckman model using the copula method, as in real life many data could not follow a BVN distribution. This facilitates the Heckman model estimation and makes it more robust to deviations from the assumptions regarding the distribution of the data.

@@

For coverage estimation, we obtain the confidence intervals from the Wald method. For the random effect variance, even when it is possible to estimate the confidence intervals through the profile or bootstrap method, we did not estimated them for computational reasons.

>>>>

---
title: Multiple imputation of incomplete multilevel data using Heckman selection models
author:
- name: J. Munoz*
  num: a
- name: V. de Jong
  num: a
- name: T. Debray
  num: a
address:
- num: a
  org: Julius Center for Health Sciences and Primary Care, Utrecht Medical Center,Utrech, The Netherlands
corres: "* Johanna Munoz. \\email{j.munozavila@umcutrecht.nl}"
presentaddress: Julius Center for Health Sciences and Primary Care, Utrecht Medical Center, Universiteitsweg 100, 3584 CG Utrech, The Netherlands
authormark: Uthor \emph{et al}.
articletype: Research article
received: 2022-11-01
revised: 2022-12-01
accepted: 2023-01-01
abstract: "Missing data is a common problem in medical research, and is commonly addressed using multiple imputation. Although traditional imputation methods allow for valid statistical inference when data are missing at random (MAR), their implementation is not justified when observations are clustered (e.g., within studies) or when the presence of missingness depends on unobserved information. Unfortunately, this situation is increasingly common, and typically arises when individual participant data (IPD) are combined from multiple studies. While several imputation methods have been proposed for addressing individual studies where data are missing not at random (MNAR), their application and validity in large datasets with clustering remains unclear. We therefore explored the consequence of MNAR data in IPDMA in-depth, and proposed novel multilevel imputation methods for common missing patterns in clustered datasets. These methods build upon the principles of Heckman selection models, and adopt a two-stage meta-analysis approach for imputing binary and continuous variables. After evaluating the proposed imputation models in simulated scenarios, we illustrated their use in a malaria cross-sectional community survey to estimate the prevalence of parasitemia for children aged 2-10 years in five subregions in Uganda."
keywords: Heckman model; IPDMA; Missing not at random; Selection models;Multiple imputation;
bibliography: bibfile.bib
output: 
  rticles::sim_article:
  fig_caption: yes
  longtable: yes
  doublespace: yes

@@

Under the BVN assumption, the parameters of the Heckman model coefficients can be estimated using the two-step Heckman method (H2S) [@heckman1976] or the full information maximum likelihood method (FIML) [@amemiya1984]. However, both methods can lead to inconsistent estimators when this underlying distribution cannot be assumed[@gomes_etal20].To overcome this problem, other approaches have been proposed that relax the distribution assumptions, among them copula models [@smith03]. The copula approach uses a function, known as copula, that joins the marginal distribution of the error terms of the selection and outcome equation which are specified separately.
$$ F(r^*_{i},y^*_{i}) =\Phi\left(r^*_{i}-X^S_{i}\beta^S_{i},\frac{y^*_{i}-X^O_{i}\beta^O_{i}}{\sigma_i}, \rho_i\right)$$
Thus, to estimate the parameters of the Heckman method, it is sufficient to specify the marginal distributions of the error terms, and link them with a suitable copula function. In our imputation method we estimate the Heckman model using the copula method, as in real life many data could not follow a BVN distribution. This facilitates the Heckman model estimation and makes it more robust to deviations from the assumptions regarding the distribution of the data.

@@

To calculate the coverage of the parameter coefficients, we estimate confidence intervals (CI) with the Wald method. Although it is possible to obtain CI through the profile or bootstrap method for the variances of the random effects, we prefer not to estimate them (and hence the coverage) for these parameters due to computational time.
